Formal Construction of Microcontroller Instruction Set Model Using B

### 2011-10-07 papers/ArticleZ80/source/paper.tex
Small update

\begin{abstract}
  [TODO:To redo]
  This paper describes an approach to model the functional aspects of
  the instruction set of microcontroller platforms using the notation
  of the B method. The paper presents specifically the case of the Z80 
  platform. This work is a contribution towards the extension of the
  B method to handle developments up to assembly level code.
\end{abstract}
%
%\keywords{Fomal Methods, microcontroller Verification, Hardware}

\section{Introduction}
[TODO:To update] 
The B method~\cite{Abrial} supports the construction of safety systems
models by verification of proofs that guarantees its correctness. So,
an initial abstract model of the system requirements is defined and
then it is refined until the implementation model. Development
environments based on the B method also include source code generators
for programming languages, but the result of this translation
cannot be compared by formal means. The paper \cite{Dantas_SBMF08} presented
recently an approach to extend the scope of the B method up to the
assembly level language. One key component of this approach
is to build, within the framework of the B method, formal models of
the instruction set of such assembly languages.

@@

Other possible uses of a formal model of a microcontroller
instruction set include documentation, the construction of simulators,
and be possibly the starting point of a verification effort for the
actual implementation of a Z80 design. Moreover the model of the
instruction set could be instrumented with non-functional aspects,
such as the number of cycles it takes to execute an instruction, to
prove lower and upper bounds on the execution time of a routine. The
goal of this project, though, is to provide a basis for the generation
of software artifacts at the assembly level that are amenable to 
refinement verification within the B method.

This paper is focused on the presentation of modelling of assembly instruction
set, including elementary libraries to describe hardware aspects. The paper is
structured as follows. Section~\ref{sec:B_method} provides a short introduction
to the B method. Section~\ref{sec:models} presents the elementary libraries and
the modelling of some elements common to microcontrollers.  Section~\ref{sec:z80}
presents the B model of the Z80 instruction set. Section ~\ref{sec:Verification}
provides some information on the proof effort needed to analyze the presented
models. Related work is discussed in Section~\ref{sec:relatedworks}. Finally, the
last section is devoted to the conclusions.

\subsection{Problems in formalizing instruction set}
[I need review the text and move this secion to other place]
%a) Explicar como sheo formalizados os conjuntos de instruheheo nos
%manuais dos construtores (Consultar os dois manuais)

In general, the manuals of microcontrollers can be more appropriate for
developers in assembly language. Many times, each instruction of microcontrollers
are shown in its official manual using textual description, math partially
description of instruction, examples, encoding, number of cycles and others
informations. But the explantion about the action of instruction do not use
usually the better way to describe the instruction for users and developers. For example,
if the user want to know the action of instruction and its effect on the flag
register then he needs search this information in different pages of manual.
Besides, the writer of manual usually describes important facts in not formal
notation and default. So this description can add some ambiguities or until
mistakes.

% b) The case of Z80
The official manual of the microcontroller Z80 \cite{Z80_manual} has several
problems that were identified over time~\cite{UndocumentedZ80}. The Z80 is a
microcontroller that  has been tested and used for many years. This intensive use
and test facilitated to find several errors in the description of the
instructions in its manual. Several of these problems have been cataloged in a
technical report \cite{UndocumentedZ80}. Some problems are inaccurate
information, for example the description of action of instruction on the flag
register, partial informations and distributed on different pages of manual.

Some erros was identified during the development of Z80 model.
An example is
%\begin{itemize}
% \item Some input and output instructionsThe
% INI/INIR/IND/INDR/OUTI/OUTD/OTIR/OTDR instructions do affect the CF flag (some official documentation says they leave it unaffected, important!) and the
% NF flag is not always set but may also be reset (see 4.3 for exact operation)
% \cite{UndocumentedZ80}.
% Este erro he descrito na phegina (274 do documento e 294 do pdf) 
% do manual oficial do Z80 (Z80_CPU_adotado.pdf)
% Citar apenas esse exemplo abaixo! 
%   \item
when a determined interruption (NMI) is accepted, the bit IFF1 is not
copied to bit IFF2. % Este erro he descrito na phegina (23 do documento e 41do pdf) % do manual oficial do Z80 (Z80_CPU_adotado.pdf)
Only IFF1 is reset. This problem was identified, because the description of
interruption on the official manual is big and complex, so to help the
construction of Z80 B model, others references  \cite{UndocumentedZ80,Simulator_z80} 
also were consulted.

%\end{itemize}

An interesting solution is to specify formally the assembly instruction set,
this creates a pattern of representation and validates properties of
instructions. The formal model also restricts the definitions to correct typing and use only
expressions well-defined.
% -> Solucao interessante he especificar  formalmente o conjunto
% de instruhehehees, isto cria um padrao formal de representaheheheo e valida propriedades
% das instrucoes. 
Furthermore, the developer can use with B method different levels of
abstraction and add more specific details in refined models.

% allow to verify properties on the model.
\section{Introduction to the B Method}
\label{sec:B_method}

@@

A B specification is structured in modules. A module defines a set of valid
states, including a set of initial states, and operations that may provoke a
transition between states. The design process starts with a module with a
so-called functional model of the system under development. In this initial
modeling stage, the B method requires that the user proves that, in a machine,
all the its initial states are valid, and that operations do not define
transitions from valid states to invalid states.

Essentially, a B module contains two main parts: a header and the available
operations. Figure~\ref{fig:maqB} has a very basic example. The clause
$\mathit{MACHINE}$ has the name of module.  The next two clauses respectively
reference external modules and create an instance of an external module. The
$\mathit{VARIABLES}$ clauses declares the name of the variables that compose the
state of the machine. Next, the $\mathit{INVARIANT}$ clause defines the type and
other restrictions on the variables. The $\mathit{INITIALIZATION} $ specifies
the initial states. Finally, operations correspond to the transitions between
states of the machine.

@@

The entities defined in the module $\mathit{BIT\_DEFINITION}$ are the
type for bits, logical operations on bits (negation, conjunction,
disjunction, exclusive disjunction), as well as a conversion function
from booleans to bits.

@@

Conjunction is an unary function on bits and it is defined as:

@@

Finally, the conversion from booleans to bits is simply defined as:

@@

Usually, each type module just needs to instantiate concepts that were already defined in the hardware
modelling library.  For example, the function $\mathit{bv\_to\_nat}$ from bit vector arithmetics is
specialized to $\mathit{byte\_uchar}$. As the set $\mathit{BYTE}$ is a subset of the
$\mathit{BIT\_VECTOR}$, this function can defined as follows:

@@

The \textit{Z80} is a CISC microcontroller developed by \textit{Zilog}~\cite{Z80_manual}. It supports 158
different instructions, including all the 78 from Intel 8080 microprocessor, and all of them were
specified. These instructions are classified into these categories: load and exchange; block transfer and
search; arithmetic and logical; rotate and shift; bit manipulation; jump, call and return; input/output;
and basic cpu control. Each category of instruction has different elements of
specification.

@@

\subsection{Modelling registers and input and output ports} [TODO:I need review
the english]

@@

\subsection{Flag register} [TODO:I need review the text]
Another important element is the ``z'' register ($\mathit{rgs8(z0)}$),  that is used as a flag register. This
register uses only six bits to represent the execution result status of each instruction.
According to the official manual the bits 3 and 5 are not used and the others bits have the follow meaning:
\begin{description}
  \item[$\mathit{bv\_get(rgs8(z0),0)}$] - The Carry bit
  \item[$\mathit{bv\_get(rgs8(z0),1)}$] - The Add/Subtract bit
  \item[$\mathit{bv\_get(rgs8(z0),2)}$] - The Parity or Overflow bit
  \item[$\mathit{bv\_get(rgs8(z0),4)}$] - The Half Carry bit
  \item[$\mathit{bv\_get(rgs8(z0),6)}$] - The Zero bit
  \item[$\mathit{bv\_get(rgs8(z0),7)}$] - The Sign bit 
\end{description}

@@

\textbf{Assuring the absence of overflow:}
 \emph{To assure that not to happen overflow, the developer can add this
 expression ($\mathit{bv\_get(rgs8(z0),0)} \neq 1 \land \mathit{bv\_get(rgs8(z0),2) \neq 1}$)
 in the invariant. By default, the overflow can happen, but many times it can be dangerous. 
 Then, the developer may prohibit its use. This restriction can also become the more difficult
 to specify.}

\subsection{Manipulation data functions from Z80}
[I need to review the text]\\ There are some specific functions
from Z80 to manipulate the data. In addressing mode, the function $\mathit{bv\_ireg\_plus\_d}$ is  used to indexed address. It receives the
value of register ($\mathit{ix}$ or $\mathit{iy}$) and the displacement to return the sum, the result is the dislocated address memory, see
its the definition.

@@

\subsection{Program, stack and data memory}[I need review the text]

The Z80 uses a unique memory for storing program instructions, data stack and data work. The memory has 16-bits
addressing and each address represents a byte. Thus the data from the memory module is very simple, as shown below,
but an additional care must be taken to preserve the consistency of memory.

@@

In general, the instructions can access all memory address, but it is very dangerous. For add more security, is
important that the program instructions has the access limited by region. Thus the designer can specify address
regions to restrict the access from instructions. The address regions can be specified how shown below.
The constants ($\mathit{PRORGRAM\_R\_ADR,DATA\_R\_ADR,STACK\_R\_ADR}$) defines respectively a range of restricted address
for: programs instructions, data work, data stack.

@@

  \item \it instruction\_next  $\in$  USHORT  $\fun$  USHORT \rm - It receives the  
  actual value from program counter register (\textit{pc}) and returns its increment.
 \item \it is\_negative  $\in$  \it BYTE  $\fun$  \it BIT \rm - It returns the most significant bit,
 in others words, the signal bit.

@@

Every logic functions that are defined in the \textit{BYTE} and \textit{BV16} module are included in the \textit{ALU}
module. Then, these functions can be seen directly in the \textit{ALU} and \textit{Z80} module.

\subsection{Modelling the instructions}[I need review the text]

Each instruction is represented by a B operation in the module Z80. The main module (\textit{Z80}) has two
types of operations, a type represents the external actions and the other represents the instructions of
microcontrollers. The external actions are shown in the section ~\ref{sec:externalactions}.  A simple
example from instruction is a $\mathit{LD\_(nn)\_A}$\footnoteremember{myfootnote}{The tools B does not allow use parentheses in identifiers,
then the characters \{``('',``)''\} are replaced respectively by \{``9'',``0''\} in the real specification.} shown below.
Many times to model a instruction is necessary to use the pre-defined functions, this facilitates the construction of model. By default, all
parameters from operations are either predefined elements in the model or integers values in the decimal
representation. This instruction use the $\mathit{updateAddressMem}$ function from \textit{Memory} module
and it receives a address memory and its new memory value. After, it increments the program counter
($\mathit{pc}$) and update the refresh register ($\mathit{r\_}$).

@@

\subsection{Modeling the input/output instructions}

@@

The $\mathit{IN\_r(C)}$\footnoterecall{myfootnote} instruction is represented
by the follow B operation. It is from I/O group, then it receives a identifier of register
$\mathit{r}$ and, in this place, it stores the value of $\mathit{C}$ port address. Besides, it increments the
program counter and updates the flag registers.

@@

The main I/O instructions work similarly, for example: $\mathit{OUT(n),A}$  or
$\mathit{OUT (C ), r}$. There are several others instructions of I/O, besides these are
the most commonly used.

@@

\subsection{Modeling the external actions}
\label{sec:externalactions}

@@

The others external actions are related to interrupts. The interrupts allows that the devices suspend a routine from CPU and start
another service routine. This service routine can exchange data or signals between CPU and external devices. When a
routine is finished, then the CPU come back to the last routine that was interrupted.

@@

The $\mathit{iff1}$ and $\mathit{iff2}$ control the maskable interrupts ($\mathit{INT}$). When the $\mathit{iff1}$
is set, the interrupt is enable, otherwise it is disable. The $\mathit{iff2}$ is used only as a tempory storage
place for $\mathit{iff1}$. The $\mathit{iff1}$ is 1 or 0 by instructions $\mathit{EI}$ and $\mathit{DI}$, that respectively
enable and disable the maskable interruptions.

The interruptions and the \textit{reset} action change the state of
microcontroller. Theses actions are modeled by B operations and the its main
effects are shown below\footnote{Some definitions of constants: $\mathit{sp\_minus\_two}$ is the value of stack pointer minus 2,
%=$\mathit{dec\_BV16(dec\_BV16(sp))}$
 $\mathit{sp\_minus\_one}$ is the value of stack pointer minus 1,
%=$\mathit{dec\_BV16(sp)}$
$\mathit{pc\_high}$ is the most significant 8 bits and
$\mathit{pc\_low}$ is the others least significant.}.

 \textbf{NMI} - Non-maskable interrupts - The non-maskable cannot be disable
 by the programmer. Then, when a device makes a request, the $sp$ is pushed, the $pc$ receive
 $66H$ (102 in decimal), the $\mathit{iff1}$ is reset , $\mathit{iff2}$ stores
 $\mathit{iff1}$ and refresh register is updated.

@@

  \textbf{INT} - Maskable Interrupt -  This is usually reserved to important functions that can be enabled and
  disabled by the programmer. When a maskable interrupt action happens, both $\mathit{iff1}$ and $\mathit{iff2}$ are
  cleared, disabling the interrupts, the $sp$ is pushed, the refresh register is updated  and the other effects
  depend on the interrupt mode.

@@

  \item The mode 0 is compatible with 8080 and  \it im \rm = \rm ( \rm 0 $\mapsto$  \rm 0 \rm ). When a 
  non-maskable interruption happen , the CPU fetches an instruction of one byte from an external device, usually an
  RST instruction, and the CPU executes it. The instruction code is received from a external device by data bus and
  it is represented by integer parameter called $\mathit{byte\_bus}$ .

@@

  \item The mode 2 is the most flexible and  \it im \rm = \rm ( \rm 1  $\mapsto$ \rm 1 \rm ). When a non-maskable
  interruption happens, an indirect call can be made to any address memory. The program counter receives a bit vector of size 16 composed in the
  part most significant the $\mathit{i\_}$ register and, in the  part least significant, the $\mathit{byte\_bus}$
  with the last bit reset.
 \end{itemize}

@@

\subsection{Conclusions about the Z80 model}

The Z80 formal model provides many benefits, because the verification of generated proof obligations assure: 
the use correct of data types, the developed security properties  and if the expressions used is well-defined(WD)\footnote{An expression
is called ``well-defined'' (or unambiguous) if its definition assigns it a unique interpretation or value.}. Furthermore, the designer has
a big flexibility to create news and specifics security properties, this is very useful to adjust the verification in accordance with the requirements.
Besides, this example of model could replace or improve the documentation used for users and assembly programmers. Z80 model is also useful to
develop verified software  until the assembly level. There are too much benefits when a formal model is developed, 
but verify completely the model is not an easy task and more details are shown in section\ref{sec:Verification}. 
Finally, the next section explain a developed study case and verified since first abstract B model until B assembly model.

@@

=\hspace*{0.10in}\rm (\rm (\it initial\_level \rm - \it final\_level\rm )  $\mod$  \rm 2\rm 5\rm 6\rm ) 

@@

\section{Verification}
\label{sec:Verification}

[To get new informations from dissertation]
% [O que as provas garantem? citar que dificeis, que he necessherio buscar thecnicas eficientes
% para realizar as provas ]

The proof obligations allow to verify the data types, important system
properties and if the expressions are well-defined (WD). The
properties provide additional guarantees, because they can set many
safety rules. However, the model can be very difficult to prove.
Some statistics are shown below:
%[Groups]
\begin{center}
\begin{tabular}{llll}
\textbf{Group} &  \textbf{Proofs No Obvious} & \textbf{Proofs WD} &   \\ 
Input and Output & 54 & 296  \\ 
Logic and Arithmetic & 134 & 413  \\ 
Manipulation of Bits & 160 & 742  \\ 
Extern actions & 51 & 84  \\
General & 140 & 615  \\
Initialization, Properties and Assertions & 68 & 169  \\
 \textbf{Total} &  607 &  2319  &  %\textbf{Total} 2926 \\
\end{tabular}
\end{center}

@@

There are in the literature of computer science some
approaches~\cite{BHDL_2003,GEMPLUS_99} to model hardware and the virtual machines
using the B method. Then, in both works the B method has been used successfully
to model the operational semantic. However the cost of modeling was still expensive and this
paper quoted some techniques to lower the cost of modeling.
% Although, the modelling of the Z80 do not have modelled some small aspects.

@@

The main motivation of our research is the development of verified
software up to the assembly level, which requires specifying the
semantics of the underlying hardware. Thus some aspects were not
modeled in our work such as the execution time of the instructions.
Also we did not consider the micro architecture of the hardware as the scope of
our work does not include hardware verification. However, there are
many other specialized techniques to verify these questions.

\section{Conclusions}
\label{sec:conclusions}

>>>>

\begin{abstract}
  This paper describes an approach to model the functional aspects of
  the instruction set of microcontroller platforms and an embedded software
  verified up to assembly level. These models were developed using the notation
  of the B method. The paper presents specifically the case of the Z80 
  platform and case study software that can be used in tanks of the petroleum
  industry. This work is a contribution towards the extension of the B method
  to handle developments up to the assembly level code.
\end{abstract}
%
%\keywords{ Validation, Verification, Microcontroller, Hardware}
\paragraph[Keywords]{ \textbf{Keywords}: Validation, Verification,  Embedded Software and
Hardware}

\section{Introduction}

The B method~\cite{Abrial} supports the construction of safety systems
models by verification of proofs that guarantee its correctness. So,
an initial abstract model of the system requirements is defined and
then it is refined up to the implementation model. Development
environments based on the B method also include source code generators
for programming languages, but the result of this translation
cannot be compared by formal means. The paper \cite{Dantas_SBMF08} has presented
recently an approach to extend the scope of the B method up to the
assembly level language. One key component of this approach
is to build, within the framework of the B method, formal models of
the instruction set of such assembly languages.

@@

Other possible uses of a formal model of a microcontroller
instruction set include documentation, the construction of simulators,
and can be possibly the starting point of a verification effort for the
actual implementation of a Z80 design. Moreover, the model of the
instruction set could be instrumented with non-functional aspects,
such as the number of cycles it takes to execute an instruction, to
prove lower and upper bounds on the execution time of a routine.

\subsection{Problems in formalizing instruction set}

In general, the manuals of microcontrollers can be more suited for
developers in assembly language. Many times, each instruction of microcontrollers
is shown in its official manual using textual description, math partially
description of instruction, examples, encoding, number of cycles and other
information. But the explanations about the action of instruction do not use
usually the best way to describe the instruction for users and developers. For
example, if the users want to know the action of instruction and its effects on
the flag register then he needs to search this information in different pages of
the manual. Besides, the writer of the manual usually describes important facts
in not formal notation and default. So, this description can add some
ambiguities or even mistakes.

% b) The case of Z80
The official manual of the microcontroller Z80 \cite{Z80_manual} has several
problems that were identified over time~\cite{UndocumentedZ80}. The Z80 is a
microcontroller that  has been tested and used for many years. This intensive use
and test facilitated to find several errors in the description of the
instructions in its manual. Several of these problems have been reported in a
technical report \cite{UndocumentedZ80}. Some problems are inaccurated
information, for example the description of action of instruction on the flag
register, partial information and distributed on different pages of manual.

A simple example of error in Z80 Manual happens
% \begin{itemize} \item Some input and output instructionsThe
% INI/INIR/IND/INDR/OUTI/OUTD/OTIR/OTDR instructions do affect the CF flag (some
% official documentation says they leave it unaffected, important!) and the NF
% flag is not always set but may also be reset (see 4.3 for exact operation)
% \cite{UndocumentedZ80}. Este erro he descrito na phegina (274 do documento e
% 294 do pdf) do manual oficial do Z80 (Z80_CPU_adotado.pdf) Citar apenas esse
% exemplo abaixo! \item
when a determined interruption (NMI) is accepted, the bit IFF1 is not
copied to bit IFF2~\cite{UndocumentedZ80}.  This example can cause serious
problems for programmers.
% Este erro he descrito na phegina (22 do documento e 26do pdf) do manual
% official do Z80 (Z80_CPU_adotado.pdf) Only IFF1 is reset.
This error was identified, because the description of interruption on the
official manual is big and complex, so to help the construction of Z80 B model,
other references \cite{UndocumentedZ80,Simulator_z80}  were also consulted.

An interesting solution is to specify formally the assembly instruction set, this
creates a pattern of representation and validates properties of instructions. The
formal model also restricts the definitions to correct typing and uses only
expressions well-defined.Furthermore, the developer can use with B method
different levels of abstraction and add more specific details in refined models.

Finally, the main goal of this project, though, is to provide a basis for the
generation of software artefacts at the assembly level, those are amenable to 
refinement verification within the B method.

\subparagraph{Report structure}

This work is focused on the presentation of modelling of assembly instruction
set, including elementary libraries to describe hardware aspects; and the
modelling and simulation of a case study of an embedded verified software that
can be used in tanks of the petroleum industry.

The report is structured as it follows. Section~\ref{sec:B_method} provides a
short introduction to the B method. Section~\ref{sec:models} presents the elementary libraries and
the modelling of some elements common to microcontrollers.  Section~\ref{sec:z80}
presents the B model of the Z80 instruction set. Section ~\ref{sec:Verification}
provides some information on the proof effort needed to analyse the presented
models. Section ~\ref{sec:studycase} explains the building and verifying
assembly-level refinements. Related works are discussed in
section~\ref{sec:relatedworks}. Finally, the last section is devoted to the conclusions.

\section{Introduction to the B Method}
\label{sec:B_method}

@@

A B specification is structured in modules. A module defines a set of valid
states, including a set of initial states, and operations that may provoke a
transition among states. The design process starts with a module with a
so-called functional model of the system under development. In this initial
modelling stage, the B method requires that the user proves that, in a machine,
all its initial states are valid, and operations do not define
transitions from valid states to invalid states.

Essentially, a B module contains two main parts: a header and the available
operations. Figure~\ref{fig:maqB} has a very basic example. The clause
$\mathit{MACHINE}$ has the name of module.  The next two clauses, respectively,
reference external modules and create an instance of an external module. The
$\mathit{VARIABLES}$ clause declares the name of the variables that compose the
state of the machine. Next, the $\mathit{INVARIANT}$ clause defines the type and
other restrictions on the variables. The $\mathit{INITIALIZATION} $ specifies
the initial states. Finally, operations correspond to the transitions among
states of the machine.

@@

The entities defined in the module $\mathit{BIT\_DEFINITION}$ are the
type for bits, logical operations on bits (negation, conjunction,
disjunction, exclusive disjunction), as well as a conversion function
from Boolean to bit.

@@

Conjunction is a unary function on bits and it is defined as:

@@

Finally, the conversion from Boolean to bit is simply defined as:

@@

Usually, each type module just needs to instantiate concepts that were already defined in the hardware
modelling library.  For example, the function $\mathit{bv\_to\_nat}$ from bit vector arithmetic is
specialized to $\mathit{byte\_uchar}$. As the set $\mathit{BYTE}$ is a subset of the
$\mathit{BIT\_VECTOR}$, this function can defined as follows:

@@

The \textit{Z80} is a CISC microcontroller developed by
\textit{Zilog}~\cite{Z80_manual}. The Z80 is composed by different elements and a
simplified internal organization is shown on the Figure \ref{fig:DiagramBlock}.
This Figure has some important elements of Z80 CPU: ALU, registers of 8 and 16
bits and input/output ports.

\begin{figure}[h] \centering
\includegraphics[width=0.78\textwidth]{images/Architecture.png}
\caption{Internal diagram block of Z80 CPU.}
\label{fig:DiagramBlock}
\end{figure}

Z80 supports 158 different instructions, including all the 78 from Intel 8080
microprocessor, and all of them were specified. These instructions are classified
into these categories: load and exchange; block transfer and search; arithmetic
and logical; rotate and shift; bit manipulation; jump, call and return;
input/output; and basic cpu control. Each category of instruction has different
elements of specification.

@@

\subsection{Modelling registers and input and output ports}

@@

\subsection{Flag register}
Another important element is the ``f'' register ($\mathit{rgs8(f0)}$),  that is
used as a flag register. This register uses only six bits to represent the execution result status of each instruction.
According to the official manual the bits 3 and 5 are not used and the others bits have the follow meaning:
\begin{description}
  \item[$\mathit{bv\_get(rgs8(f0),0)}$] - The Carry bit
  \item[$\mathit{bv\_get(rgs8(f0),1)}$] - The Add/Subtract bit
  \item[$\mathit{bv\_get(rgs8(f0),2)}$] - The Parity or Overflow bit
  \item[$\mathit{bv\_get(rgs8(f0),4)}$] - The Half Carry bit
  \item[$\mathit{bv\_get(rgs8(f0),6)}$] - The Zero bit
  \item[$\mathit{bv\_get(rgs8(f0),7)}$] - The Sign bit 
\end{description}

@@

\textbf{Assuring the absence of overflow:}
 \emph{To assure that an overflow does not happen, the developer can add this
 expression ($\mathit{bv\_get(rgs8(f0),0)} \neq 1 \land
 \mathit{bv\_get(rgs8(f0),2) \neq 1}$) in the invariant. By default, the overflow
 can happen, but it can be dangerous many times. Then, the developer may
 prohibit its use. This restriction can also become more difficult to specify
 the instructions.}

\subsection{Manipulation data functions from Z80}
 There are some specific functions
from Z80 to manipulate the data. In addressing mode, the function
$\mathit{bv\_ireg\_plus\_d}$ is  used to indexed address. It receives the value
of register ($\mathit{ix}$ or $\mathit{iy}$) and the displacement to return the
sum, the result is the dislocated address memory, see its definition.

@@

\subsection{Program, stack and data memory}

The Z80 uses a unique memory for storing program instructions, data stack and
data work. The memory has 16-bits address and each address represents a byte.
Thus, the data from the memory module is very simple, as shown below, but an
additional care must be taken to preserve the memory consistency.

@@

In general, the instructions can access all memory address, but it is very
dangerous. For adding more security, it is important that the program
instructions has the limited access by region. Thus the designer can specify
address regions to restrict the access from instructions. The address regions can
be specified how shown below. The constants
($\mathit{PRORGRAM\_R\_ADR,DATA\_R\_ADR,STACK\_R\_ADR}$) defines respectively a
range of restricted address for: programs instructions, data work, data stack.

@@

  \item \it instruction\_next  $\in$  USHORT  $\fun$  USHORT \rm - It receives the  
  actual value from program counter register (\textit{pc}) and returns its increment.
 \item \it is\_negative  $\in$  \it BYTE  $\fun$  \it BIT \rm - It returns the most significant bit,
 in other words, the signal bit.

@@

The logic functions that are defined in the \textit{BYTE} and \textit{BV16}
module are included in the \textit{ALU} module. Then, these functions can be seen
directly in the \textit{ALU} and \textit{Z80} module.

\subsection{Modelling the instructions}

Each instruction is represented by a B operation in the module Z80. The main
module (\textit{Z80}) has two types of operations, a type represents the external
actions and the other represents the instructions of microcontrollers. The
external actions are shown in the section ~\ref{sec:externalactions}.  A simple
example from instruction is a
$\mathit{LD\_(nn)\_A}$\footnoteremember{myfootnote}{The tools B does not allow
to use parentheses in identifiers, then the characters \{``('',``)''\} are
replaced respectively by \{``9'',``0''\} in the real specification.} shown below. The
pre-defined functions is necessary many times to model the instructions, this
functions facilitates the construction of instruction set model. By default, all
parameters from operations are either predefined elements in the model or
integers values in the decimal representation. This instruction use the
$\mathit{updateAddressMem}$ function from \textit{Memory} module and it receives
a address memory and its new memory value. After, it increments the program
counter ($\mathit{pc}$) and update the refresh register ($\mathit{r\_}$).

@@

\subsection{Modelling the input/output instructions}

@@

The $\mathit{IN\_r(C)}$\footnoterecall{myfootnote} instruction is represented
by the following B operation. It is from I/O group, then it receives an
identifier of register $\mathit{r}$ and, in this place, it stores the value of $\mathit{C}$ port address. Besides, it increments the
program counter and updates the flag registers.

@@

The main I/O instructions work similarly, for example: $\mathit{OUT(n),A}$  or
$\mathit{OUT (C ), r}$. There are several other instructions of I/O, besides,
these ones are the most commonly used.

@@

\subsection{Modelling the external actions}
\label{sec:externalactions}

@@

The other external actions are related to interruptions. The interruptions allow
that the devices suspend a routine from CPU and start another service routine.
This service routine can exchange data or signals between CPU and external
devices. When a routine is finished, then the CPU comes back to the last routine
that was interrupted.

@@

The $\mathit{iff1}$ and $\mathit{iff2}$ control the maskable interrupts
($\mathit{INT}$). When the $\mathit{iff1}$ is set, the interrupt is enabled,
otherwise it is disabled. The $\mathit{iff2}$ is used only as a temporay storage
place for $\mathit{iff1}$. The $\mathit{iff1}$ is 1 or 0 by instructions
$\mathit{EI}$ and $\mathit{DI}$, that respectively enable and disable the
maskable interruptions.

The interruptions and the \textit{reset} action change the state of
microcontroller. Theses actions are modelled by B operations and the its main
effects are shown below\footnote{Some definitions of constants: $\mathit{sp\_minus\_two}$ is the value of stack pointer minus 2,
%=$\mathit{dec\_BV16(dec\_BV16(sp))}$
 $\mathit{sp\_minus\_one}$ is the value of stack pointer minus 1,
%=$\mathit{dec\_BV16(sp)}$
$\mathit{pc\_high}$ is the most significant 8 bits and
$\mathit{pc\_low}$ is the others least significant.}.

 \textbf{NMI} - Non-maskable interrupts - The non-maskable cannot be disabled
 by the programmer. Then, when a device makes a request, the $sp$ is pushed, the
 $pc$ receives $66H$ (102 in decimal), the $\mathit{iff1}$ is reset , $\mathit{iff2}$ stores
 $\mathit{iff1}$ and refresh register is updated.

@@

  \textbf{INT} - Maskable Interrupt -  This is usually reserved to important functions that can be enabled and
  disabled by the programmer. When a maskable interrupt action happens, both $\mathit{iff1}$ and $\mathit{iff2}$ are
  cleared, disabling the interrupts, the $sp$ is pushed, the refresh register is updated  and the other effects
  depend on the interrupt mode (\it im \rm).

@@

  \item The mode 0 is compatible with 8080 and  \it im \rm = \rm ( \rm 0 $\mapsto$  \rm 0 \rm ). When a 
  non-maskable interruption happens, the CPU fetches an instruction of one byte
  from an external device, usually an RST instruction, and the CPU executes it.
  The instruction code is received from an external device by data bus and it is represented by integer parameter called $\mathit{byte\_bus}$ .

@@

  \item The mode 2 is the most flexible and  \it im \rm = \rm ( \rm 1  $\mapsto$
  \rm 1 \rm ). When a non-maskable interruption happens, an indirect call can be
  made to any address memory. The program counter receives a bit vector of size
  16 composed with two parts: the most significant part of the $\mathit{i\_}$
  register and the least significant part of the $\mathit{byte\_bus}$ with the
  last bit cleared.
 \end{itemize}

@@

\subsection{Considerations about the Z80 model}

The Z80 formal model provides many benefits, because of the verification of
generated proof obligations assure: the correct use of data types, the developed
security properties  and if the expressions used are well-defined
(WD)\footnote{An expression is called ``well-defined'' (or unambiguous) if its definition assigns
it a unique interpretation or value.}. Furthermore, the designer has a big
flexibility to create new and specific security properties, this is very useful
to adjust the verification in accordance with the requirements. Moreover, this
example of model could replace or improve the used documentation for users and
assembly programmers. Z80 model is also useful to develop verified software 
up to the assembly level. There are many benefits when a formal model is
developed, but to verify completely the model is not an easy task and more
details are shown in section~\ref{sec:Verification}. Finally, the next section explains
a verified case study since the first abstract B model up to B
assembly model.

@@

\hspace*{1.10in}\it byte\_uchar\rm (\it io\_ports\rm (\it uchar\_byte\rm (\rm
2\rm )\rm )\rm ) \rm =\hspace*{0.10in}\rm (\rm (\it initial\_level \rm - \it
final\_level\rm )  $\mod$  \rm 2\rm 5\rm 6\rm )\\ \hspace*{1.10in} $\land$\\

@@

\section{Verification Process}
\label{sec:Verification}

% [O que as provas garantem? citar que dificeis, que he necessherio buscar thecnicas eficientes
% para realizar as provas ]
This section describes the verification process of B model of instructions set
and B model in assembly-level. The proof obligations allow to verify the data
types, important system properties and if the expressions are well-defined (WD). The
properties provide additional guarantees, because they can set many
safety rules. However, the model can be very difficult to prove without using
different strategies.

The use of proof commands can accelerate the verification process, therefore two
proof commands are simple, effective, efficient and useful; they are presented
in~\ref{subsection:VerificationInstructions}. When these two commands are used is
recommended to set the time out of theorem prover to one second, which is enough
to solve the vast majority of proof obligations and prevent the theorem prover
waste time unnecessarily in proof obligations that are not resolved by these
proof commands. The next two subsections present these two proof commands that
solve quickly the most of proof obligations.

% O uso de comandos de prova pode acelerar bastante o processo de prova, portanto
% dois comandos de provas simples,  eficazes, eficientes e muito úteis são
% apresentados. Quando esses dois comandos de provas são utilizados é recomendado
% ajustar o \textit{time out} do provador para 1 segundo, o que é suficiente para
% resolver a grande maioria das obrigações de prova e isso evita que o provador
% desperdice muito tempo desnecessariamente em obrigações de prova que não são
% resolvidas por estes comandos.

The verification process is organized in groups two groups: instructions set
verification and assembly-level verification.

\subsection{Verification process of instructions set model} 
\label{subsection:VerificationInstructions}
The following proof command is used to solve the most of WD proof obligatios of
Z80 model. This proof command is applied in proof obligations that satisfy the pattern: $x \in dom(y)$.
The sequence of commands: select the minimum force in theorem prover, remove the obvious hypothesis from stack, 
replace the term $dom(y)$ for an equivalent expression, execute the simplifier
and the automatic prover.

% O comando de prova seguinte é utilizado para resolver a maioria das obrigações de prova EBD do modelo do
% Z80. Esse comando de prova é aplicado nas obrigações de prova que satisfazem o padrão: $x \in dom(y)$.
% A sequência de comandos: seleciona a força mínima do provador, remove as hipóteses óbvias da pilha,
% substitui o termo $dom(y)$ por um equivalente, executa o simplificador e o provador automático.

\begin{center}
\framebox(350,70)[c]{ Pattern( x $\in$ dom(y) ) \& ff(0) \& dd \& eh(dom(y)) \& ss \& pr }
\end{center}

Some statistics proofs of instructions set model are shown below:
%[Groups]
\begin{center}
\begin{tabular}{llll}
\textbf{Instructions Groups} &  \textbf{Proofs No Obvious} & \textbf{Proofs WD}
& \\ Input and Output & 54 & 296  \\ 
Logic and Arithmetic & 134 & 413  \\ 
Manipulation of Bits & 160 & 742  \\ 
Extern actions & 51 & 84  \\
General & 140 & 615  \\
Initialization, Properties and Assertions & 68 & 169  \\
 \textbf{Total} &  607 &  2319  &  %\textbf{Total} 2926 \\
\end{tabular}
\end{center}

@@

There are in the literature of computer science some
approaches~\cite{BHDL_2003,GEMPLUS_99,DBLP:journals/entcs/EvansG08,DBLP:conf/fmco/Leuschel08,DBLP:conf/asm/Wright08}
to model hardware and the virtual machines using B. Then, in these works the B
method has been used successfully to model the operational semantic. However the
cost of modelling was still expensive and this report quoted some techniques to
lower the cost of modelling.
% Although, the modelling of the Z80 do not have modelled some small aspects.

@@

The main motivation of our research is the development of verified
software up to the assembly level, which requires specifying the
semantics of the underlying hardware. Thus some aspects were not
modelled in our work such as the execution time of the instructions.
Also we did not consider the micro architecture of the hardware as the scope of
our work does not include hardware verification. However, there are
many other specialized techniques to verify these questions.
